SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2020-01-21 2017-Xudong-SMT/ActaInformatica.tex
Acta paper: More explanation in Sect.4.2 and 4.3.

We first define the set of actions a PLTS can use.  Let $a$
range over action constants, $\symb{op}$ over operators, and $x$  over
variable names. Action terms are given by the following grammar:
\[
\begin{array}[l]{rcl@{\quad}p{5.5cm}}
\alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
p_i&::=& ~\symb{Expr}&\text{parameters}\\
\symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,\dots,\symb{Expr}_n)&\text{expressions}
\end{array}
\]
%\[
%\begin{array}[l]{rcl@{\quad}p{5.5cm}}
%  \alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
%  p_i&::=& ?x~|~\symb{Expr}&\text{parameters (input variable or expression)}\\
%  \symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,..,\symb{Expr}_n)&\text{Expressions}
%\end{array}
%\]

@@

Before the SAT checking step, the set $L$ may contain symbolic transitions that
do not have valid ground instantiations. This is not logically
incorrect, but is \QINX{clearly} not an optimal representation. So (line 11) we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and then checked for
satisfiability (function \textit{filterUnsatTransitions}). 
Then we compute (lines 12-14) the set of reachable transitions and states to build
the final open automaton (more details in Section~\ref{section:BuildingTheOpenAutomaton}).

@@

In Algorithm~\ref{alg1}, functions \textit{filterUnsatTransitions} (in line 11) 
\textit{refineReachableSubAutomaton} (in line 13) make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) values of
variable $v$.

\ERIC{Each call to the SMT engine will return either SAT or UNSAT results, or may fail due to time bounds. Also, the accuracy of a ``SAT'' result may depend on the characteristics of the specific theory used to axiomatise the data domains and operators. Such theories may be intrinsically undecidable, or may not  be properly handled by the SMT engine (e.g. for recursive datatypes).
However, while building the open automaton, which is a symbolic representation of the pNet behaviour, it is not logically incorrect to keep potentially unsatisfiable transitions, so we will treat undecidable results, and time bounds failures, as SAT. So the open automaton may have more transitions and states than the theoretical minimum.}

@@

A further reduction may now be gained by using the information from
these sets of possible assignments: adding this information when
checking satisfiability of transitions leaving a given state may
result in more UNSAT transitions. The set of possible
assignments computed by the algorithm is an over approximation of
possible values of the 
variables of such state, so narrowing it means we get a better
approximation, and potentially more information, which can be used in
the SAT checking of subsequent transitions. In turn, this may reduce the possible
assignments of the corresponding target states. In
Algorithm~\ref{alg1}, we iterate this procedure (lines 13-14), each iteration
reducing the set of 
transitions and the sets of assignments; the final open automaton
is the fixed point of this iteration.

In practice, this could be
expensive \QIN{as it requires several traversals of the whole set of open transitions}, and it is not incorrect to leave some UNSAT OTs in the
result. \QIN{\YX{Empirically, on real life systems, the unreachable transitions constitute  a small part of the ultimate} open transitions, very little will be gained to compute the fixed point. So we would only perform one step of iteration.} Further algorithms 
(e.g. model-checking or equivalence checking) will be able to use the
full information from the predicates and the assignment sets, even if
some spurious OTs are left here.

\begin{algorithm}[htbp]
  \caption{\QINX{makeReachableSubAutomaton()}}
  \label{alg:makeReachableSubAutomaton}
\begin{algorithmic}[1]
\Require \QINX{The initial state of the pNet $\textit{InitState}(P)$. A set of satisfiable transitions $L$.}
\State Initialize a set $A$;
\State Store $(\textit{InitState}(P), \textit{InitState}(P).Assignments)$ into $S$;  \hfill $\setminus \setminus$$S.Assignments$ is the assignment of the variable in the state $S$.
\While{!$\mathit{isEmpty}(A)$}
	\For {each $(S, Assignments) \in A$}
          \For {each $OT$ from $S$}
          \State Check the predicate of $OT$ taking $Assignments$ in to account;
	  \IfThenElse{$\textit{false}$}{\textbf{skip};}{Add target of $OT$ and its assignments ($S'$, $S'.Assignments$) into $A$;}
	\EndFor	\EndFor
\EndWhile
\State Build $\mathit{OA}$ with the set $L$;
\State \QINX{\Return $\mathit{OA}$;}
\end{algorithmic}  
\end{algorithm}

\begin{algorithm}[htbp]
  \caption{\QINX{refineReachableSubAutomaton()}}
  \label{alg:refineReachableSubAutomaton}
\begin{algorithmic}[1]
\Require \QINX{An open automaton $\mathit{OA}$.}
\Repeat
\For {\QINX{each global state $S$ in $\mathit{OA}$}}
          \State \QINX{$b=$ whether $S$ has preceeding transitions.}
	  \IfThenElse{$b$}{\textbf{skip};}{\QINX{remove $S$ and the transitions from $S$;}}
\EndFor
\Until  \QINX{$\mathit{OA}$ does not change;}
\State \QINX{\Return $\mathit{OA}$;}
\end{algorithmic}  
\end{algorithm}

>>>>

We first define the set of actions a PLTS can use.  Let $a$
range over action constants, $\symb{op}$ over operators, and $x$  over
variable names. Action terms (parameterised actions) are given by the following grammar:
\[
\begin{array}[l]{rcl@{\quad}p{5.5cm}}
\alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
p_i&::=& ~\symb{Expr}&\text{parameters}\\
\symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,\dots,\symb{Expr}_n)&\text{expressions}
\end{array}
\]
%\[
%\begin{array}[l]{rcl@{\quad}p{5.5cm}}
%  \alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
%  p_i&::=& ?x~|~\symb{Expr}&\text{parameters (input variable or expression)}\\
%  \symb{Expr}&::=& \symb{Value}~|~x~|~\symb{op}(\symb{Expr}_1,..,\symb{Expr}_n)&\text{Expressions}
%\end{array}
%\]

@@

Before the SAT checking step, the set $L$ may contain symbolic transitions that
do not have valid ground instantiations. This is not logically
incorrect, but is \QINX{clearly} not an optimal representation. So (line 11) we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions (see Section~\ref{section:pruning}), and then checked for
satisfiability (function \textit{filterUnsatTransitions}). 
\QINX{Then starting at the initial state, we compute (lines 12) the set of reachable transitions and states taking the assignments into account. We also iteratively prune (lines 13-14) the unreachable states by refining the possible set of values of states variables (more details in Section~\ref{section:BuildingTheOpenAutomaton}).}

@@

In Algorithm~\ref{alg1}, functions \textit{filterUnsatTransitions} (in line 11), 
\textit{refineReachableSubAutomaton} (in line 13) make use of the Z3 engine to implement the
satisfiability checking, for all the OTs computed that far.
For a transition
\[		\inferrule*[myfraction=\reddottedrule]
		{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
			\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
		{s \xrightarrow {\alpha}s'}
                \]
                we need to build a predicate encoding both the OT predicate and
the information available on the possible assignments of the initial
state $s$. This is expressed as
\[Pred_{l}\wedge \bigwedge_{v\in vars(s)}\{ \bigvee_{Assign\in
  assigns(v,s)} Assign\}
\]
where each $Assign$ represents one of the possible (symbolic) values of
variable $v$.

\ERIC{Each call to the SMT engine will return either SAT or UNSAT results, or may fail due to time bounds. Also, the accuracy of a ``SAT'' result may depend on the characteristics of the specific theory used to axiomatise the data domains and operators. Such theories may be intrinsically undecidable, or may not  be properly handled by the SMT engine (e.g. for recursive datatypes).
\QINX{When building the open automaton, which is a symbolic representation of the pNet behaviour, a symbolic transition can be pruned if its predicate is checked unsatisfiable. However in some cases SMT finds it undecidable, so we may still have some spurious symbolic transitions in the result. We cannot pretend to get minimality, but still it is better to prune as many unsatisfiable transitions as possible, together with some states which become unreachable. So we will treat undecidable results, and time bounds failures, as SAT.} So the open automaton may have more transitions and states than the theoretical minimum.}

@@

\QINX{A further reduction may now be gained by using the information from
these sets of possible assignments: adding this information when
checking satisfiability of transitions leaving a given state may
result in more UNSAT transitions. Furthermore, if there are assignments of state variables inside some loops of the automaton, and the set of their possible values may influence some predicates, we will need more iterations. Because in that case more predicates of transitions will be unsatisfiable after pruning some assignments in the previous iteration.} The set of possible
assignments computed by the algorithm is an over approximation of
possible values of the 
variables of such state, so narrowing it means we get a better
approximation, and potentially more information, which can be used in
the SAT checking of subsequent transitions. In turn, this may reduce the possible
assignments of the corresponding target states. In
Algorithm~\ref{alg1}, we iterate this procedure (lines 13-14), each iteration
reducing the set of 
transitions and the sets of assignments; the final open automaton
is the fixed point of this iteration.

In practice, this could be
expensive as it requires several traversals of the whole set of open transitions, and it is not incorrect to leave some UNSAT OTs in the
result. Empirically, on real life systems, the unreachable transitions constitute  a small part of the ultimate open transitions, very little will be gained to compute the fixed point. So we would only perform one step of iteration. Further algorithms 
(e.g. model-checking or equivalence checking) will be able to use the
full information from the predicates and the assignment sets, even if
some spurious OTs are left here.

%\begin{algorithm}[htbp]
%  \caption{\QINX{makeReachableSubAutomaton()}}
%  \label{alg:makeReachableSubAutomaton}
%\begin{algorithmic}[1]
%\Require \QINX{The initial state of the pNet $\textit{InitState}(P)$. A set of satisfiable transitions $L$.}
%\State Initialize a set $A$;
%\State Store $(\textit{InitState}(P), \textit{InitState}(P).Assignments)$ into $S$;  \hfill $\setminus \setminus$$S.Assignments$ is the assignment of the variable in the state $S$.
%\While{!$\mathit{isEmpty}(A)$}
%	\For {each $(S, Assignments) \in A$}
%          \For {each $OT$ from $S$}
%          \State Check the predicate of $OT$ taking $Assignments$ in to account;
%	  \IfThenElse{$\textit{false}$}{\textbf{skip};}{Add target of $OT$ and its assignments ($S'$, $S'.Assignments$) into $A$;}
%	\EndFor	\EndFor
%\EndWhile
%\State Build $\mathit{OA}$ with the set $L$;
%\State \QINX{\Return $\mathit{OA}$;}
%\end{algorithmic}  
%\end{algorithm}
%
%\begin{algorithm}[htbp]
%  \caption{\QINX{refineReachableSubAutomaton()}}
%  \label{alg:refineReachableSubAutomaton}
%\begin{algorithmic}[1]
%\Require \QINX{An open automaton $\mathit{OA}$.}
%\Repeat
%\For {\QINX{each global state $S$ in $\mathit{OA}$}}
%          \State \QINX{$b=$ whether $S$ has preceeding transitions.}
%	  \IfThenElse{$b$}{\textbf{skip};}{\QINX{remove $S$ and the transitions from $S$;}}
%\EndFor
%\Until  \QINX{$\mathit{OA}$ does not change;}
%\State \QINX{\Return $\mathit{OA}$;}
%\end{algorithmic}  
%\end{algorithm}
